Deep recurrent neural networks for supernovae classification

### 2017-01-31 paper/sn_submission_v3/Referee reply.txt

(1) The work as presented is in almost acceptable form. The part about using just the
first few points is powerful and will be welcomed by many. If using the entire
light-curve during testing provides somewhat worse results, then perhaps there
is a way to improve the model even further.

There is possibly ways to improve the model further, but this is not certain. One reason that improving the model further may not actually help is that, due to the way RNNs work, including the information from after the early epoch challenge skews the overall weights of the network meaning that the early epoch weights are less well known compared to training using only the early epoch challenge data. Since this is mentioned in the text we have not expanded further.

@@

(2) Many machine learning methods seem like black-boxes and keeping them that way
does not help dispel myths about them from the eyes of the uninitiated. When
describing example data, can some stats be provided about the amount of
chunking in light-curves? For example the range of time-span for the
light-curves, the number of chunks they are typically divided in (a box-plot
could be used to describe median, show outliers etc.) Or point to some
specifics from the SPCC figures (table X, Fig Y etc.)

We have added (new) Fig.2 which contains three subplots. The lower left subplot is simply the total number of days of observation of the light-curve vs. the number of grouped-time data vector elements these are split into. This shows that when there are more days of observation the light-curve data is grouped into longer data vectors. The upper subplot shows the distribution of the total number of days of observation of the light-curve with min, max, mean and median values indicated and the lower right subplot shows the distribution of the number of elements in the grouped-time data vector. We have added this information to the text.

@@

(3) *ANY* discussion about error-bars will be helpful in letting readers decide if
they would want to apply the technique. [See comment below about 73+/-2%]. But
thats just a suggestion and not a requirement.

The errors in the results are due to training 5 times with different random selections of training data (and random augmentations of data). This is mentioned in the results section already, but we have expanded on it in the same paragraph.

@@

(4) Did the authors try removing known points, then apply their chunking and random
filling and see if they get consistent results?

We ran the training removing 10% of the known points from each filter before grouping the times and randomly augmenting. The equivalent of the top line in the table with 10% of the points removed gives results very close to the results in the second line of the table, i.e. there is a slight degredation but this is to be expected, and the results are possibly better than we expected when performing this proceduce. We have added this to the results section.

@@

(5) The discussion about padding is unclear. if, in the end, the padding does not
get used, may be the discussion should be removed to an appendix or a footnote
(but do not remove it as it is part of the flow). Perhaps an example will help?

The padding is used to create the vectors of the correct length. This padding is then ignored by the network using a technique called masking. This means the padding has no effect on the results, it just maintains the correct length vectors. This is already explicitly described in the text. We feel this short description helps with the understanding of the process of training the network, but does not need expanding.

@@

(6a) In Table 2 please include the 0.052% even if the results are bad.

They are already in the table in every case - 1,045 is equivalent to a train fraction of 0.052 (not 0.052%).

@@

(6b) Are the early-epoch results for SN1a or 123? Can the results be provided for both sets?

The early-epoch challenge is for SN1a only and not 123. We have given 123 a go and added the results to the table. The difference between 123 categorisation using the full light-curve and the early-epoch data is very similar to the difference between the SN1a full light-curve and early-epoch data. The results are not as good, but still surprisingly high considering the length of the vectors.

@@

(6c) In the middle section, in the last 2 lines, Completeness seems to rise when Host
information is dropped. Thats anomalous. If you understand why, a comment will
be useful. Otherwise those are the kind of error-bars that become meaningful:
are both 73+/-2%? You can still fit the table on a single page by abbreviating
contents in the first column.

This was a statistical fluctuation. We have added the errors on the purity and completeness. As it can be seen the errors on both of the bottom lines of the second section are ~4%, approximately the same as the difference between the means. The same is true for the 123 categorisation in the early-epoch challenge where the completeness without host-z is higher than with host-z, but the error on the completeness is also very large.

@@

(7) It is understandable that your methods are not comparable with various methods
used during SPCC, but perhaps its still worth mentioning the best numbers
there?

To do...

(8) The early-epoch result is surprising. May be it should be emphasized more? It
will be useful to many people.

We have highlighted the result in the abstract as well as separated the paragraph in the conclusion section. The main point of the paper is to describe the network architecture and it's use in classification, rather than achieve the best results possible. It is somewhat fortunate that we did as well as we did on the early-epoch challenge, but we don't want to make that the main focus of the paper.

@@

(9) The following is a loose sentence: "Moreover, we have ...trained on the whole
light-curve". Especially so since you do not normalize time to the peak value so the
starting point could be arbitrary (or is your padding mitigating exactly that?
if so, clarify a little). Otherwise quantify (at least verbally) what you mean
by the above sentence.

We mean that training using the data from the early epoch challenge gives a better early-time predictor than training using the whole light-curve. We have rewritten the sentence to make it clearer.

(10) You mention that perhaps even raw images could be used etc., but you have not
even discussed the effects of different filters etc. What happens if you use
just one or two filters, for instance? So, either go into more details, or
exclude pure speculation.

This isn't pure speculation, this is a consideration for further work. It is possible to data mine images and pipe this into an RNN. We could do this for a single filter or we could do this for many filters. If one filter was missing then the input could be augmented as done here and the other raw images convolved to extract features. The principal is identical to what is done here, but with a huge amount more information. We have not changed the text here as it is an interesting prospect which others (as well as ourselves) might like to consider.

@@

(11) Section 4: Over-fitting can also show up as some inputs being learnt too well
(and thus its not the noise that gets fitted)

We have changed this to describe that over-fitting occurs when a network learns about relations between the inputs and outputs in the training data which aren't present in the test data.

@@

(12) You mention in the comments that the code is public. Please refer to it so that
it can be used by others.

This is already mentioned, the link is on the front page.

(13) Shorten the y-axes in figure 3 (keeping it same for both parts) to reduce
unused white-space.

We have changed the scale to show between 3e-2 and 4e-1 therefore reducing the unused white-space.

### 2017-01-31 paper/sn_submission_v3/ms.tex
Paper update

\begin{abstract}
   We apply deep recurrent neural networks, which are capable of learning complex sequential information, to classify supernovae\footnote{Code available at \href{https://github.com/adammoss/supernovae}{https://github.com/adammoss/supernovae}}. The observational time and  filter fluxes are used as inputs to the network, but since the inputs are agnostic additional data such as host galaxy information can also be included. Using the Supernovae Photometric Classification Challenge (SPCC) data, we find that deep networks are capable of learning about light curves, however the performance of the network is highly sensitive to the amount of training data.  For a training size of 50\% of the representational SPCC dataset (around $10^4$ supernovae) we obtain a type-Ia vs. non-type-Ia classification accuracy of 94.8\%, an area under the Receiver Operating Characteristic curve AUC of 0.986 and a SPCC figure-of-merit $F_1=0.64$. We also apply a pre-trained model to obtain classification probabilities as a function of time, and show it can give early indications of supernovae type. Our method is competitive with existing algorithms and has applications for future large-scale photometric surveys. 
   \end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{f1.pdf}
\caption{\label{fig:network} Bidirectional recurrent neural network for sequence classification. The input vectors at each sequential step are fed into a pair of bidirectional hidden layers, which can propagate information forwards and backwards. These are then merged to obtain a consensus view of the network, and finally a softmax layer computes classification probabilities. 
 }
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{f2.pdf}
\caption{\label{fig:lightcurve} (Top) Example light curve in the 4 $g,r,i,z$ bands for SN ID 551675 (a type-Ia) in the Supernovae Photometric Classification Challenge data~\cite{Kessler:2010wk}. The data has been processed using augmentation so there is a $g,r,i,z$ value at each sequential step. (Bottom) Type-Ia probability as a function of time from a 2 layer LSTM model, trained with around $10^4$ supernovae and SN 551675 excluded. The final probability gives $99.5\%$ confidence that the supernovae is of type-Ia. 
 }
\end{figure}

@@

The goal of the classifier is to determine the supernovae type in the test set. We consider two problems, (1) to categorise two classes (type-Ia vs. non-type-Ia), and (2) to categorise three classes (supernovae types-1, -2 and -3). We denote these as  `SN1a' and  `123' respectively. We also attempt to categorise between type-Ia and non-type-Ia supernovae using only the first six observations with $S/N>4$ and the data taken on the night of the sixth observation as described in~\cite{Kessler:2010wk}.

@@

\begin{figure}
\centering
\includegraphics[width=88mm, angle=0]{f3.pdf}
\caption{\label{fig:loss} (Left) Training loss (green) vs. test loss (blue) for a unidirectional 2 layer LSTM network with 16 hidden units in each layer. (Right) Training accuracy (green) vs. test accuracy (blue) for the same network.  }
\end{figure}

A dataset of 21319 is relatively small by deep learning standards. Furthermore, the `feature space' of supernovae light curves is significantly smaller than, say, using RNNs to learn about language. We therefore need to be careful about {\em over-fitting}. Over-fitting arises when the network fits noise between the inputs and outputs of the training data, which do not exist in the test data. It can typically be detected by comparing the loss of the training and test data. If the loss of training data continues to decrease, but the loss of the test data increases, this is a sure sign of over-fitting. If no sign of over-fitting is observed, the network is not usually complex enough to fully learn the relationship between inputs and outputs (called {\em under-fitting}).

@@

In Fig.~\ref{fig:loss} we show the training and tests losses for such a network, with a dropout of $0.5$, applied to type-Ia vs. non-type-Ia classification with host galaxy photo-z information. Without dropout the training loss continues to fall and the test loss rises. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $94.9 \pm 0.2$\%, AUC of $0.986 \pm 0.001$ and $F_1 = 0.64 \pm 0.01$. The corresponding type-Ia purity and completeness are $87.3 \pm 0.8\%$ and $91.4\pm 1.1\%$ respectively. A summary of results and comparisons can be found in Table~\ref{tab:05_SN1a}. The inclusion of host galaxy photo-$z$ marginally improves the classifier performance.

\begin{table*}[t!]
\centering
\begin{tabular}{ c  c  c  c  c  c  c  c }\hline\hline
Method & Training size & AUC & Accuracy (\%) & $F_1$ & Purity (\%) & Completeness (\%) & Host $z$ \\\hline\hline
A & 10,660 & $0.986 \pm 0.001$ & $94.7 \pm 0.2$ & $0.64 \pm 0.01$ & $87.3 \pm 0.8$ & $91.4 \pm 1.1$ & True \\
A & 10,660 & $0.981 \pm 0.001$ & $93.6 \pm 0.3$ & $0.60 \pm 0.02$ & $87.4 \pm 1.7$ & $85.4 \pm 2.6$ & False \\
A & 5,330 & $0.975 \pm 0.003$ & $92.9 \pm 0.6$ & $0.57 \pm 0.03$ & $86.6\pm 2.0$ & $83.4 \pm 3.4$ & True \\
A & 5,330 & $0.973 \pm 0.002$ & $92.3 \pm 0.4$ & $0.55 \pm 0.02$ & $86.2 \pm 2.4$ & $80.8 \pm 3.8$ & False \\
B & 1,103 & $0.910 \pm 0.012$ & $85.9 \pm 0.9$ & $0.31 \pm 0.03$ & $72.4 \pm 0.4$ & $66.1 \pm 6.0$ & True \\
B & 1,103 & $0.901 \pm 0.016$ & $84.6 \pm 1.7$ & $0.28 \pm 0.05$ & $68.2 \pm 3.4$ & $66.3 \pm 5.5$ & False \\
C & $\sim$10,660 & - & - & 0.58 & 85 & 88 & True \\
C & $\sim$10,660 & - & - & 0.51 & 82 & 85 & False \\
C & 1,045 & - & - & 0.33 & 70 & 75 & True \\
C & 1,045 & - & - & 0.29 & 67 & 71 & False \\
D & $\sim$8,000 & - & - & 0.55 & - & - & True \\
D & $\sim$2,00$\underset{~}{0}$ & - & - & 0.45 & - & - & True \\
\hline\hline
E & 10,660 & $0.974 \pm 0.001$ & $90.4 \pm 0.3$ & - & $90.6 \pm 0.7$ & $86.5 \pm 0.7$ & True\\
E & 10,660 & $0.959 \pm 0.006$ & $88.5 \pm 1.1$ & - & $87.6 \pm 1.1$ & $85.9 \pm 4.1$ & False\\
F & 1,103 & $0.868 \pm 0.015$ & $78.1 \pm 0.9$ & - & $70.8 \pm 3.4$ & $70.6 \pm 4.1$ & True\\
F & 1,103 & $0.865 \pm 0.011$ & $78.0 \pm 1.2$ & - & $66.9 \pm 3.2$ & $74.5 \pm 4.2$ & False\\\hline\hline
A & 10,660 & $0.977 \pm 0.002$ & $93.1 \pm 0.4$ & $0.58 \pm 0.01$ & $88.0 \pm 1.1$ & $82.2 \pm 2.8$ & True\\
A & 10,660 & $0.970 \pm 0.001$ & $92.0 \pm 0.3$ & $0.53 \pm 0.01$ & $86.0 \pm 0.9$ & $79.5 \pm 2.2$ & False\\
B & 1,103 & $0.902 \pm 0.014$ & $85.2 \pm 1.2$ & $0.29 \pm 0.04$ & $71.5 \pm 1.6$ & $62.8 \pm 5.6$ & True\\
B & 1,103 & $0.860 \pm 0.017$ & $81.6 \pm 1.2$ & $0.21 \pm 0.02$ & $62.6 \pm 3.0$ & $57.6 \pm 2.7$ & False\\\hline\hline
E & 10,660 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & True\\
E & 10,660 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & False\\
F & 1,103 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & True\\
F & 1,103 & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & $*** \pm ***$ & False\\\hline\hline
\end{tabular}
\caption{\label{tab:05_SN1a} (Top section) Summary of results for  type-Ia vs. non-type-Ia classification with a training fraction of 0.5, 0.25 and 0.052 with comparisons to similar methods in~\cite{Karpenka:2012pm} and~\cite{Newling:2010bp}. (Second section) Summary of results for types-I,~-II~and~-III classification. (Third section) Summary of results for SPCC early-epoch challenge. (Bottom section) Summary of the results for the SPCC early-epoch challenge when classifying between types-I~,~-II~and~-III supernovae. The models used are A) Unidirectional LSTM, [16, 16] with 0.5 dropout, B) Unidirectional LSTM, [4] with 0.5 dropout, C) \cite{Karpenka:2012pm}, D) \cite{Newling:2010bp}, E) Bidirectional LSTM, [16, 16] with 0.5 dropout, F) Bidirectional LSTM, [4] with 0.5 dropout. Errors on results are the mean and standard deviation values from 5 randomised runs.}
\end{table*}

@@

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2-layer LSTM network with 16 hidden units and dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2 class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with severe over-fitting and an increasing cross-entropy loss at the expense of metrics such as $F_1$, and decrease the model complexity.

@@

In the middle section of Table~\ref{tab:05_SN1a} the three class categorisation is shown. There is no available data for comparison of this problem, but compared to classification between type-Ia vs. non-type-Ia, bidirectional recurrent neural networks do well. The AUC and accuracy remain high, still above 90\% when the host-$z$ is included using a training fraction of 0.5

Finally the bottom section of Table~\ref{tab:05_SN1a} shows the results of the early-epoch challenge from SPCC. Here only the data before the night of the sixth observation with $S/N>4$ for each light-curve can be used - a great reduction from the use of the full light-curve. We do surprisingly well in this case obtaining an accuracy of $93.1\pm0.4\%$, AUC of $0.977\pm0.002$ and an $F_1=0.58\pm 0.01$ with a training fraction of 0.5 and including host-$z$. These values are not far from those obtained using the whole light-curve and are equivalent to the full results of~\cite{Karpenka:2012pm}. The results are not as good with a training fraction of 0.052, but still comparable to our results using the whole light-curve. The network trained on the partial light-curves does better than suggested from feeding the early-epoch light-curve through a network trained on the full sequence. This is due to the later parts of the light-curve influencing the weights of the network whilst training. Training on only the initial part of the light-curve optimises the network weights such that early sequence features have more effect, resulting in better accuracy, AUC and $F_1$ values than expected.

@@

As well as finding competitive results for the final metrics, we have shown that it is possible to give fast, early evaluation of supernovae type using pre-trained models. This is possible since the light curve can be fed to the model directly without needing any feature extraction. Moreover, we have found that training a network only on the early epoch light-curve data results in a better early-time predictor than using a network trained on entire light-curve data.

>>>>

\begin{abstract}
   We apply deep recurrent neural networks, which are capable of learning complex sequential information, to classify supernovae\footnote{Code available at \href{https://github.com/adammoss/supernovae}{https://github.com/adammoss/supernovae}}. The observational time and filter fluxes are used as inputs to the network, but since the inputs are agnostic additional data such as host galaxy information can also be included. Using the Supernovae Photometric Classification Challenge (SPCC) data, we find that deep networks are capable of learning about light curves, however the performance of the network is highly sensitive to the amount of training data.  For a training size of 50\% of the representational SPCC dataset (around $10^4$ supernovae) we obtain a type-Ia vs. non-type-Ia classification accuracy of 94.7\%, an area under the Receiver Operating Characteristic curve AUC of 0.986 and a SPCC figure-of-merit $F_1=0.64$. When using only the data for the early-epoch challenge defined by the SPCC we achieve a classification accuracy of 93.1\%, AUC of 0.977 and $F_1=0.58$, results almost as good as with the whole light-curve. By employing bidirectional neural networks we can acquire impressive classification results between supernovae types -I,~-II and~-III at an accuracy of 90.4\% and AUC of 0.974. We also apply a pre-trained model to obtain classification probabilities as a function of time, and show it can give early indications of supernovae type. Our method is competitive with existing algorithms and has applications for future large-scale photometric surveys. 
   \end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=83mm, angle=0]{f1.pdf}
\caption{\label{fig:network} Bidirectional recurrent neural network for sequence classification. The input vectors at each sequential step are fed into a pair of bidirectional hidden layers, which can propagate information forwards and backwards. These are then merged to obtain a consensus view of the network, and finally a softmax layer computes classification probabilities. 
 }
\end{figure}

@@

In Fig.~\ref{fig:violinplot} we show how the length of the grouped-time data vector is related to the duration of the light-curve. The bottom left subplot shows that more total number of day since the beginning of observation of the light-curve results in a greater number of grouped time elements in the vector. The upper subplot shows the distribution of observation lengths in the SPCC data varies significantly with two distinct peaks. These are grouped into an average of 40-3element data vectors as can be seen in the bottom right subplot.

\begin{figure}
\centering
\includegraphics[width=84mm, angle=0]{f2.pdf}
\caption{\label{fig:violinplot}. (Top) Distribution of the total number of days for each light-curve with the minimum, maximum, mean and median values indicated. (Bottom right) Distribution of the number of elements in the grouped time vector with the minimum, maximum, mean and median values indicated. (Bottom left) The trend showing that more days in the light-curve result in longer group time vectors.
 }
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=84mm, angle=0]{f3.pdf}
\caption{\label{fig:lightcurve} (Top) Example light curve in the 4 $g,r,i,z$ bands for SN ID 551675 (a type-Ia) in the Supernovae Photometric Classification Challenge data~\cite{Kessler:2010wk}. The data has been processed using augmentation so there is a $g,r,i,z$ value at each sequential step. (Bottom) Type-Ia probability as a function of time from a 2 layer LSTM model, trained with around $10^4$ supernovae and SN 551675 excluded. The final probability gives $99.5\%$ confidence that the supernovae is of type-Ia. 
 }
\end{figure}

@@

The goal of the classifier is to determine the supernovae type in the test set. We consider two problems, (1) to categorise two classes (type-Ia vs. non-type-Ia), and (2) to categorise three classes (supernovae types-1, -2 and -3). We denote these as  `SN1a' and  `123' respectively. We also attempt the first two problems using only the first six observations with $S/N>4$ and the data taken on the night of the sixth observation as described in~\cite{Kessler:2010wk}.

@@

\begin{figure}
\centering
\includegraphics[width=84mm, angle=0]{f4.pdf}
\caption{\label{fig:loss} (Left) Training loss (green) vs. test loss (blue) for a unidirectional 2 layer LSTM network with 16 hidden units in each layer. (Right) Training accuracy (green) vs. test accuracy (blue) for the same network.  }
\end{figure}

A dataset of 21319 is relatively small by deep learning standards. Furthermore, the `feature space' of supernovae light curves is significantly smaller than, say, using RNNs to learn about language. We therefore need to be careful about {\em over-fitting}. Over-fitting arises when the network learns about relations between the inputs and outputs of the training data, which do not exist in the test data. It can typically be detected by comparing the loss of the training and test data. If the loss of training data continues to decrease, but the loss of the test data increases, this is a sure sign of over-fitting. If no sign of over-fitting is observed, the network is not usually complex enough to fully learn the relationship between inputs and outputs (called {\em under-fitting}).

@@

In Fig.~\ref{fig:loss} we show the training and tests losses for such a network, with a dropout of $0.5$, applied to type-Ia vs. non-type-Ia classification with host galaxy photo-z information. Without dropout the training loss continues to fall and the test loss rises. For 5 randomised runs, training for 200 epochs, we obtain a classification accuracy of $94.9 \pm 0.2$\%, AUC of $0.986 \pm 0.001$ and $F_1 = 0.64 \pm 0.01$. The corresponding type-Ia purity and completeness are $87.3 \pm 0.8\%$ and $91.4\pm 1.1\%$ respectively. A summary of results and comparisons can be found in Table~\ref{tab:05_SN1a}. The inclusion of host galaxy photo-$z$ marginally improves the classifier performance. The 1$\sigma$ errors quoted in the table are the result of 5 runs where the training data is randomly chosen (and so different) each time. Some random choice of the set of light-curves are more effective for training the network than others, but it is extremely difficult to optimise this.

To test the robustness of the time-grouping method we remove 10\% of the known filter values (and/or their errors) before grouping the data into a single vector and randomly augmenting the missing values. After training we find there is a small degradation in the results, i.e. for a training fraction of 0.5 using a deep 2-layer, unidirectional network with 16 hidden units, a dropout of 0.5 and including the photo-$z$ information the obtained results are very similar to the second line in Table~\ref{tab:05_SN1a}. This shows that a reduction in 10\% of the points is similar to the omission of the photo-$z$ data and therefore the data augmentation method is extremely robust.

\begin{table*}[t!]
\centering
\begin{tabular}{ c  c  c  c  c  c  c  c }\hline\hline
Method & Training size & AUC & Accuracy (\%) & $F_1$ & Purity (\%) & Completeness (\%) & Host $z$ \\\hline\hline
A & 10,660 & $0.986 \pm 0.001$ & $94.7 \pm 0.2$ & $0.64 \pm 0.01$ & $87.3 \pm 0.8$ & $91.4 \pm 1.1$ & True \\
A & 10,660 & $0.981 \pm 0.001$ & $93.6 \pm 0.3$ & $0.60 \pm 0.02$ & $87.4 \pm 1.7$ & $85.4 \pm 2.6$ & False \\
A & 5,330 & $0.975 \pm 0.003$ & $92.9 \pm 0.6$ & $0.57 \pm 0.03$ & $86.6\pm 2.0$ & $83.4 \pm 3.4$ & True \\
A & 5,330 & $0.973 \pm 0.002$ & $92.3 \pm 0.4$ & $0.55 \pm 0.02$ & $86.2 \pm 2.4$ & $80.8 \pm 3.8$ & False \\
B & 1,103 & $0.910 \pm 0.012$ & $85.9 \pm 0.9$ & $0.31 \pm 0.03$ & $72.4 \pm 0.4$ & $66.1 \pm 6.0$ & True \\
B & 1,103 & $0.901 \pm 0.016$ & $84.6 \pm 1.7$ & $0.28 \pm 0.05$ & $68.2 \pm 3.4$ & $66.3 \pm 5.5$ & False \\
C & $\sim$10,660 & - & - & 0.58 & 85 & 88 & True \\
C & $\sim$10,660 & - & - & 0.51 & 82 & 85 & False \\
C & 1,045 & - & - & 0.33 & 70 & 75 & True \\
C & 1,045 & - & - & 0.29 & 67 & 71 & False \\
D & $\sim$8,000 & - & - & 0.55 & - & - & True \\
D & $\sim$2,00$\underset{~}{0}$ & - & - & 0.45 & - & - & True \\
\hline\hline
E & 10,660 & $0.974 \pm 0.001$ & $90.4 \pm 0.3$ & - & $90.6 \pm 0.7$ & $86.5 \pm 0.7$ & True\\
E & 10,660 & $0.959 \pm 0.006$ & $88.5 \pm 1.1$ & - & $87.6 \pm 1.1$ & $85.9 \pm 4.1$ & False\\
F & 1,103 & $0.868 \pm 0.015$ & $78.1 \pm 0.9$ & - & $70.8 \pm 3.4$ & $70.6 \pm 4.1$ & True\\
F & 1,103 & $0.865 \pm 0.011$ & $78.0 \pm 1.2$ & - & $66.9 \pm 3.2$ & $74.5 \pm 4.2$ & False\\\hline\hline
A & 10,660 & $0.977 \pm 0.002$ & $93.1 \pm 0.4$ & $0.58 \pm 0.01$ & $88.0 \pm 1.1$ & $82.2 \pm 2.8$ & True\\
A & 10,660 & $0.970 \pm 0.001$ & $92.0 \pm 0.3$ & $0.53 \pm 0.01$ & $86.0 \pm 0.9$ & $79.5 \pm 2.2$ & False\\
B & 1,103 & $0.902 \pm 0.014$ & $85.2 \pm 1.2$ & $0.29 \pm 0.04$ & $71.5 \pm 1.6$ & $62.8 \pm 5.6$ & True\\
B & 1,103 & $0.860 \pm 0.017$ & $81.6 \pm 1.2$ & $0.21 \pm 0.02$ & $62.6 \pm 3.0$ & $57.6 \pm 2.7$ & False\\\hline\hline
A & 10,660 & $0.960 \pm 0.006$ & $87.9 \pm 0.9$ & - & $86.4 \pm 0.8$ & $84.4 \pm 3.5$ & True\\
A & 10,660 & $0.948 \pm 0.002$ & $86.8 \pm 0.3$ & - & $84.1 \pm 1.1$ & $83.7 \pm 1.4$ & False\\
B & 1,103 & $0.851 \pm 0.013$ & $76.8 \pm 1.3$ & - & $64.7 \pm 3.8$ & $71.0 \pm 4.1$ & True\\
B & 1,103 & $0.819 \pm 0.010$ & $74.2 \pm 1.0$ & - & $58.1 \pm 3.8$ & $73.6 \pm 6.6$ & False\\\hline\hline
\end{tabular}
\caption{\label{tab:05_SN1a} (Top section) Summary of results for  type-Ia vs. non-type-Ia classification with a training fraction of 0.5, 0.25 and 0.052 with comparisons to similar methods in~\cite{Karpenka:2012pm} and~\cite{Newling:2010bp}. (Second section) Summary of results for types-I,~-II~and~-III classification. (Third section) Summary of results for SPCC early-epoch challenge. (Bottom section) Summary of the results for the SPCC early-epoch challenge when classifying between types-I~,~-II~and~-III supernovae. The models used are A) Unidirectional LSTM, [16, 16] with 0.5 dropout, B) Unidirectional LSTM, [4] with 0.5 dropout, C) \cite{Karpenka:2012pm}, D) \cite{Newling:2010bp}, E) Bidirectional LSTM, [16, 16] with 0.5 dropout, F) Bidirectional LSTM, [4] with 0.5 dropout. Errors on results are the mean and standard deviation values from 5 randomised runs.}
\end{table*}

@@

For $5.2\%$ of the representative SPCC data, the training dataset is so small that over-fitting is more severe. Using the same 2-layer LSTM network with 16 hidden units and dropout of 0.5 we find a notable increase in the test loss after $\sim 20$ epochs, but the accuracy and other metrics remain relatively constant ($F_1$ values of 0.35 to 0.4 were obtained). The reason for this apparent discrepancy is that the accuracy, say, simply takes the maximum value of the softmax output layer. For example, a 2-class problem with output probabilities [0.6, 0.4] and target [1, 0] has the same accuracy as one with output probabilities [0.8, 0.2]. The loss in the latter case would be lower however, and represents increased confidence of the network in its predictions. We therefore reject models with severe over-fitting and an increasing cross-entropy loss at the expense of metrics such as $F_1$, and decrease the model complexity.

@@

In the second section of Table~\ref{tab:05_SN1a} the three class categorisation is shown. There is no available data for comparison of this problem, but compared to classification between type-Ia vs. non-type-Ia, bidirectional recurrent neural networks do well. The AUC and accuracy remain high, still above 90\% when the host-$z$ is included using a training fraction of 0.5. Using a smaller training fraction of 0.052, the results are worsened similar to the two class categorisation in the top section of Table~\ref{tab:05_SN1a}.

The third section of Table~\ref{tab:05_SN1a} shows the results of the early-epoch challenge from SPCC. Here only the data before the night of the sixth observation with $S/N>4$ for each light-curve can be used - a great reduction from the use of the full light-curve. We do surprisingly well in this case obtaining an accuracy of $93.1\pm0.4\%$, AUC of $0.977\pm0.002$ and an $F_1=0.58\pm 0.01$ with a training fraction of 0.5 and including host-$z$. These values are not far from those obtained using the whole light-curve and are equivalent to the full results of~\cite{Karpenka:2012pm}. The results are not as good with a training fraction of 0.052, but still comparable to our results using the whole light-curve. The network trained on the partial light-curves does better than suggested from feeding the early-epoch light-curve through a network trained on the full sequence. This is due to the later parts of the light-curve influencing the weights of the network whilst training. Training on only the initial part of the light-curve optimises the network weights such that early sequence features have more effect, resulting in better accuracy, AUC and $F_1$ values than expected.

Finally, the bottom section of Table~\ref{tab:05_SN1a} has the results of the three class categorisation when using the early-epoch data. The results are similar to the difference between the full light-curve and early-epoch data SN1a categorisation when comparing with the full light-curve 123 categorisation. It should be noted that the bidirectional network used for the 123 categorisation using the full light-curve revealed sizeable over-fitting when using the early-epoch data and so a unidirectional network was used instead.

@@

As well as finding competitive results for the final metrics, we have shown that it is possible to give fast, early evaluation of supernovae type using pre-trained models. This is possible since the light curve can be fed to the model directly without needing any feature extraction.

Most interestingly, we have found that training a network only on the early epoch light-curve data results in a better early-time predictor than using a network trained on entire light-curve data. Our results using only the early-epoch data are close to those using the entire light-curve data for both SN1a and 123 categorisation with both large and small training fractions.
